Investigating Neolithic pottery usage from Guijiabao in the Yanyuan Basin in Southwest China using organic residue analysis

### 2023-01-23 analysis/paper/paper.Rmd
worked on the background section

---
title: "Investigating Neolithic pottery usage in the Yanyuan Basin in Southwest China using organic residue analysis"
author:
  - Myfirstname Mylastname:
      email: fl@oneplace.edu
      institute: [UofO]
      correspondence: true
  - Yourfirstname Yourlastname:
      email: fl@another.edu
      institute: [UofA]
      correspondence: false
institute:
  - UofO: University of One Place
  - UofA: University of Another Place
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

<!-- sites and research questions-->
As an early Neolithic site, Guijiabao (ca. 4800-4300 BP) is crucial to understanding cultural interactions and the spread of crops into the Yanyuan Basin due to its location at the intersection of the middle Hengduan Mountain ranges, the Sichuan Basin, and the Yunnan-Guizhou Plateau. The recent evidence at Guijiabao reveals millet as the main crop with little rice according to the results of flotation [@citation], different from the abundant rice grains found at other Neolithic sites in that area. Although the plant remains suggest the possible domestication of millet in the early Neolithic, we still lack direct evidence of culinary practices during the Neolithic transition. To explore the extent of acceptance of new crops and species in the Yanyuan Basin, we ask the questions: What was the culinary practices during the Neolithic transition? Could we identify the evidence of new crops or species incorporated into the local dietary systems in the Early Neolithic period? We answer those questions by examining potsherds collected from Guijiabao using organic residue analysis, which could offer a better understanding of food sources during the Neolithic transition in the Yanyuan Basin.

@@

<!--Info about Yanyuan area-->
The Yanyuan area in Sichuan province is a pull-apart basin located in the middle eastern part of the Hengduan Mountain, which is a group of parallel mountain ranges in a north-south direction from the Tibetan Plateau to the Yunnan-Guizhou Plateau (Figure \@ref(fig:SW-china-map)). Although the ranges isolate areas with their own unique geographic conditions and cultures, river valleys provide pathways for the movements of people and the exchange of goods between cultural groups since the prehistoric period [@Lin2022; @Hao2022]. Therefore, the archaeological sites in the Yanyuan Basin are important to understanding the cultural interactions between southwest China and other regions, including northwest China, the Sichuan area, and northwest Yunnan.

<!--Info about Guijiabao-->
Among those archaeological sites in Yanyuan, Guijiabao located in the southwestern part of the basin is a unique site with relatively complete and continuous occupation from the early Neolithic period to the historical period [@Guijiabao2017a; @Guijiabao2017b]. Investigation in 2015 and subsequent excavations from 2016 to 2019 reveal abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guidiabao is a settlement site with long-term occupation [@Hao2022]. The chronology of Guijiabao can be divided into three temporal phases based on 19 radiocarbon dates from plant remains and a human bone, including the Neolithic period (5000-3700 BP.), the Bronze Age (3200-2700 BP.), and the late North Song to early Ming Dynasties (1200-500 BP.) [@Hao2022]. For the Neolithic period, it can be further separated into two subphases, 5000-4500 cal. BP and 4500-3700 cal. BP, according to the artifact assemblages and ceramic forms and styles. The archaeobotanical remains found at Guidiabao reveal the evidence of crop cultivation dominated by foxtail and broomcorn millet can be traced back to 5000 B.P and mixed cultivation of millet and rice in a later period (3000–1700 B.C) [@Guijiabao2014; @Hao2022].

<!--Info about Gaoshan-->
Gaoshan site is located in the southwestern part of Chengdu Plain.

@@

net sinkers
1000 BC horse
Gaushan-90 pig, plant-most are rice: 95% and a few millet

@@

>>>>

---
title: "Investigating Neolithic pottery usage from Guijiabao in the Yanyuan Basin in Southwest China using organic residue analysis"
author:
  - Myfirstname Mylastname:
      email: fl@oneplace.edu
      institute: [UofO]
      correspondence: true
  - Yourfirstname Yourlastname:
      email: fl@another.edu
      institute: [UofA]
      correspondence: false
institute:
  - UofO: University of One Place
  - UofA: University of Another Place
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

<!-- sites and research questions-->
As an early Neolithic site, Guijiabao (ca. 4800-4300 BP) is crucial to understanding cultural interactions and the spread of crops into the Yanyuan Basin due to its location at the intersection of the middle Hengduan Mountain ranges, the Sichuan Basin, and the Yunnan-Guizhou Plateau. The recent evidence at Guijiabao reveals millet as the main crop with little rice according to the results of flotation [@Guijiabao2016], different from the abundant rice grains found at other Neolithic sites in that area. Although the plant remains suggest the possible domestication of millet in the early Neolithic, we still lack direct evidence of culinary practices during the Neolithic transition. To explore the extent of acceptance of new crops and species in the Yanyuan Basin, we ask the questions: What was the culinary practices during the Neolithic transition? Could we identify the evidence of new crops or species incorporated into the local dietary systems in the Early Neolithic period? We answer those questions by examining potsherds collected from Guijiabao using organic residue analysis, which could offer a better understanding of food sources during the Neolithic transition in the Yanyuan Basin.

@@

<!--Yanyuan area-->
The Yanyuan area in Sichuan Province is a pull-apart basin located in the middle eastern part of the Hengduan Mountain, which is a group of parallel mountain ranges in a north-south direction from the Tibetan Plateau to the Yunnan-Guizhou Plateau (Figure \@ref(fig:SW-china-map)). Although the ranges isolate areas with their own unique geographic conditions and cultures, river valleys provide pathways for the movements of people and the exchange of goods between cultural groups since the prehistoric period [@Lin2022; @Hao2022]. Therefore, the archaeological sites in the Yanyuan Basin are important to understanding the cultural interactions between southwest China and other regions, including the Sichuan basin, northwest China, and northwest Yunnan.

<!--Guijiabao site-->
Among those archaeological sites in Yanyuan, Guijiabao located in the southwestern part of the basin is a unique site with relatively complete and continuous occupation from the early Neolithic period to the historical period [@Guijiabao2017a; @Guijiabao2017b]. A survey in 2015 and subsequent excavations from 2016 to 2019 reveal abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guidiabao is a settlement site with long-term occupation [@Hao2022]. The chronology of Guijiabao can be divided into three temporal phases, including the Neolithic period (ca. 5000-3700 BP), the Bronze Age (ca. 3200-2700 BP), and the late North Song to early Ming Dynasties (ca. 1200-500 BP) based on 19 radiocarbon dates from plant remains and a human bone [@Hao2022]. For the Neolithic period, it can be further separated into two subphases, 5000-4500 ca. BP and 4500-3700 ca. BP, according to the content of artifact assemblages and ceramic forms and styles.

<!--the major finding at GJB-->
Among the various ceramic styles, ceramics with surface decorations such as incised geometric patterns and impressed dot patterns are the two common styles that have also been found in western Yunnan, indicative of frequent interactions between Guijiabao and sites in Yunnan [@Guijiabao2017a]. The archaeobotanical remains reveal evidence of crop cultivation dominated by foxtail and broomcorn millet based on the analysis of flotation samples from ash pits and stratigraphic layers in 2015 [@Guijiabao2016]. The adoption of millet cultivation can be traced back to ca. 5000 BP [@Hao2022]. In addition, rice grains have also been found in ash pits throughout different phases where the earliest rice dates to 2881-2633 ca. BP [@Hao2022]. It is suggested that mixed farming of millet and rice could spread to southwest Sichuan around 5,000 years ago and to western Yunnan before 4600 years ago based on the evidence of rice remains from Guijiabao and other sites in the neighboring areas, Baiyangcun and Henglanshan [@Jiang2015; @YunnanMuseum1981; @Hao2022; @Chi2010].

<!--info about Gaoshan-->
Gaoshan site is located in the southwestern margin of the Chengdu Plain, an alluvial plain between the Sichuan Basin and the Tibetan Plateau (Figure \@ref(fig:SW-china-map)). Excavations at Gaoshan revealed abundant pottery, stone artifacts, animal and plant remains, and features such as house foundations, walls, burials, ash pits, ditches, and a well [@Gaoshan2017; @He2020; @JiangGaoshan2017; @ZhouGaoshan2015]. According to the pottery forms and decorations, Gaoshan can be categorized into the earliest phase of the Baodun period, the earliest Neolithic culture (ca. 4700-3700 BP) in the Chengdu Plain [@Guedes2013]. Gaoshan could be even older than the Baodun culture according to the burial context, which was found in the stratigraphic layer earlier than the Baodun period [@Gaoshan2017].

A large quantity of carbonized rice grains was found at Gaoshan with a relatively small number of millets, suggesting rice is the main crop supplemented by millets [@JiangGaoshan2017]. This finding also indicates the spread of rice cultivation from the Middle and Lower Yangzi valley to Sichuan in southwest China [@Chi2010]. Moreover, recent research on animal bones in 2015 and 2016 shows that pigs and dogs were domesticated, of which pigs are the main domestic animal [@He2020]. The primary subsistence strategy at Gaoshan could be rice and millet cultivation with domesticated pigs.

@@

net sinkers
1000 BC horse
Gaushan-90 pig, plant-most are rice: 95% and a few millet  
ecological zones

@@

More systematic sampling of potsherds across different phases and pottery types may eliminate a clear picture of diet and subsistence during the Neolithic transition.
